Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-04-03 2024pipit/expand-ecoop24-submit0.tex
diff

To be confident that our proofs hold on the real system, we need a single shared semantics for the compiler and the prover.
In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
To verify programs, Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics.
To execute programs, Pipit can generate executable code, which is total and semantics-preserving.

@@

\begin{itemize}
  \item we motivate the need to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit; after checking deferred properties, programs are \emph{blessed}, which marks their properties as valid (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:core:causality}, \autoref{s:transition:proof});
  \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
  \item finally, we evaluate Pipit by implementing the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver and verifying an abstract model of a key component (\autoref{s:evaluation}).
\end{itemize}

@@

To introduce Pipit, we consider a \emph{time-triggered} network driver, which has a static schedule dictating the network traffic, and which all nodes on the network must adhere to.
This driver is a simplification of the Time-Triggered Controller Area Network (TTCAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.

At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
Each basic cycle consists of a sequence of actions to be performed at specific time-marks.
Actions in the schedule may not be relevant to all nodes; the node's \emph{node matrix} contains only the relevant actions.
The node matrix is represented in memory by a \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
Watch alerts are generally placed after an expected reference message to signal an error if no reference message is received.

\autoref{f:tt-systemmatrix-ok} (left) shows an example node matrix for a non-leader node.
The matrix consists of two basic cycles C0 and C1 with messages sent at time-marks 0, 1 and 2.
The node expects to receive a reference message at time-mark 7; the watch at time-mark 9 allows a grace period before triggering an error if the reference message is not received.
\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding triggers array.

\begin{figure}
  \begin{minipage}{0.5\textwidth}
\resizebox{\textwidth}{!}{
\begin{tabular}{r|ccccc}
   & TM0 & TM1 & TM2 & $\cdots$ & TM9 \\
  \hline
  C0 & SEND A & SEND B & - & $\cdots$ & WATCH \\
  C1 & SEND A & - & SEND C & $\cdots$ & WATCH
\end{tabular}}
\end{minipage}
\begin{minipage}{0.49\textwidth}
\scriptsize
\begin{verbatim}
 0:{ time = 0; enabled = {C0,C1}; action = SEND(A); }
 1:{ time = 1; enabled = {C0};    action = SEND(B); }
 2:{ time = 2; enabled = {C1};    action = SEND(C); }
 3:{ time = 9; enabled = {C0,C1}; action = WATCH;   }
\end{verbatim}
\end{minipage}

\caption{Left: node matrix; right: corresponding triggers array configuration}
\label{f:tt-systemmatrix-ok}
\end{figure}

The network has strict timing requirements which prohibit the driver from looping through the entire triggers array at each time-mark.
Instead, the driver maintains an index that refers to the current trigger.
At each time-mark, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.

@@

We implement a streaming function \emph{count_when} to maintain the index into the triggers array; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
At each step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the counter as-is.

@@

\subsection{Restrictions on the triggers array}

Our driver may fall behind when trying to execute certain schedules, as the driver only processes one trigger per time-mark.
To ensure that the schedule can be executed on time, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next enabled trigger starts.

Recall our concrete triggers array from \autoref{f:tt-systemmatrix-ok}, which contained trigger 1 (SEND B at time-mark 1 on cycle C0), and trigger 2 (SEND C at time-mark 2 on cycle C1).
We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
However, we \emph{cannot} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.

We impose three restrictions on \emph{valid} triggers arrays: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index, so that it is reachable in time from the start of the array.

@@

We can now implement the trigger-fetch logic, which keeps track of the current trigger.
We use the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
We simplify our presentation here and only consider a constant cycle: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.

\begin{tabbing}
  \tt{MM}\= \tt{MM} \= \tt{let index} \= \kill
  \tt{let} trigger_fetch ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
    \> $\tt{rec } \textit{index}.$ \\
    \> \> $\tt{let } \textit{inc} = \text{false} \tt{ fby } ((\text{time_mark}~\textit{index}) \le \textit{time} ~\vee~ \neg (\text{enabled}~\textit{index}~\textit{cycle})) \tt{ in}$\\
    \> \> $\tt{let } \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} \tt{ in}$ \\
    \> \> $\text{pose}~(\text{lemma_can_reach_next}~\textit{cycle}~\textit{index})$; \\
    \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
    \> \> $\textit{index}$
\end{tabbing}

@@

We use the $\textit{pose}$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it.
We then state an invariant as a deferred property.
Informally, the invariant states that, either the current active trigger is not late, or the next active trigger after the current index is in the future and we can reach it in time.

With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
The contract we state is that if we are called once per time-mark then we guarantee that we never encounter a late trigger.

\begin{tabbing}
  \tt{MM}\= \tt{MM} \= \tt{MMMMMMMMMMMMM} \= \kill
  \tt{let} $\text{trigger_fetch}_{\PSValid}$ ($\textit{cycle}$: $\NN$): $\stream \NN \to \stream \NN$ = \\
  \> \tt{let} $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
  \> \> \tt{rely} = $(\lambda \textit{time}.~ \textit{time} = \xfby{0}{(\textit{time} + 1)});$ \\
  \> \> \tt{guar} = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
  \> \> \> $\implies (\text{time_mark}~\textit{index}) \ge \textit{time});$ \\
  \> \> \tt{body} = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cycle}~\textit{time} );$ \\
  \> \} \tt{in} \\
  \> \tt{assert} $(\text{Contract.inductive_check}~\textit{contract})$ \tt{by} (pipit_simplify ()); \\
  \> $\text{Contract.stream_of_contract}_1~\textit{contract}$
\end{tabbing}

@@

The key distinction between our streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous \emph{tick} of computation.
Expressing such a rely in an imperative setting would require extra encoding, as preconditions in imperative languages do not generally have an innate notion of the previous value with respect to a global shared clock.

When \emph{trigger_fetch} is used in other parts of the program, the caller must ensure that the environment satisfies the rely clause.
In the core language, this is tracked by another deferred property status attached to the contract; we will discuss this further in \autoref{s:core}.

\section{Pipit core language}
\label{s:core}

\begin{figure}
  \includegraphics[width=\textwidth]{figures/core-structure-1920x450.pdf}
\caption{Architecture of Pipit. The gray boxes and solid arrows are defined in this paper. The white boxes and dashed arrows are trusted components. The labels denote verified properties of the translation: abstraction~($\subseteq$), entailment of proof obligations~($\vdash$), and equivalence~($=$).
}
\label{f:core:structure}
\end{figure}

@@

\autoref{f:core:structure} shows the high-level architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
The translation to abstract systems also generates proof obligations, which are verified to correspond to the proof obligations on the original program.
The translation to executable transition systems is proven to be semantics-preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.

@@

\begin{figure}
  \[
  \begin{array}{lrlr}
    e, e' & := & v ~|~ x ~|~ p(\ov{e}) & \mbox{(values, variables and operations)} \\
          & | & \xfby{v}{e} ~|~ \xrec{x}{e[x]} & \mbox{(delayed and recursive streams)} \\ & | & \xlet{x}{e}{e'[x]} & \mbox{(let-expressions)}\\
          & | & \xcheckP{\PStatus}{e_{\text{prop}}} & \mbox{(checked properties)} \\
          & | & \xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}{\eguar[x]}} & \mbox{(rely-guarantee contracts)} \\
\\
    v & := & n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ r \in \mathbb{R} ~|~ \hdots  & \mbox{(values)} \\
    p & := & (+) ~|~ (-) ~|~ (\times) ~|~ \tt{if-then-else} ~|~ \hdots & \mbox{(primitives)} \\
    \\
    \PStatus & := & \PSValid ~|~ \PSUnknown & \mbox{(property statuses: valid or unknown)}\\
    \\
    V & := & \cdot ~|~ V;v & \mbox{(streams of values)} \\
    \sigma & := & \sgl{\ov{x \mapsto v}} & \mbox{(heaps)} \\
    \Sigma & := & \cdot ~|~ \Sigma;\sigma & \mbox{(streaming history environments)} \\
\tau, \tau' & := & \mathbb{N} ~|~ \mathbb{B} ~|~ \tau \times \tau ~|~ \hdots & \mbox{(value types)} \\
    \Gamma & := & \cdot ~|~ x : \tau, \Gamma & \mbox{(type environments)}  \\
    \end{array}
  \]
  \caption{Core grammar: expressions $e$, values $v$, primitive operations $p$, and property statuses $\PStatus$.}
  \label{f:core-grammar}
\end{figure} 
The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.

Recursive streams are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but the stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

@@

Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ allow modular reasoning by replacing the implementation with an abstract specification.
Contracts involve two verification conditions.
Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
Conceptually, then, a contract could have two property statuses: one for the definition and one for the instantiation.
However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers \emph{ints} by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.

The remaining grammatical constructs of \autoref{f:core-grammar} describe streams, value environments, types and type environments.
Streams $V$ are represented as a sequence of values; streaming history environments $\Sigma$ are streams of heaps.
Types $\tau$ and type environments $\Gamma$ are standard.
For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.

@@

  \caption{Typing rules for Pipit; the judgment $\typing{\Gamma}{e}{\tau}$ denotes that expression $e$ describes a \emph{stream} of values of type $\tau$. Auxiliary functions are used for values and primitive operations.}\label{f:core-typing}
\end{figure} 
We define the typing judgments for Pipit in \autoref{f:core-typing}.
Most of the typing rules are standard for an unclocked Lustre.
The typing judgment $\typing{\Gamma}{e}{\tau}$ denotes that, in an environment of streams $\Gamma$, expression $e$ denotes a stream of type $\tau$.
This core typing judgment differs from the surface syntax used in \autoref{s:motivation}, which used an explicit stream type; for the core language, we instead assume that everything is a stream.

We use an auxiliary function $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$; for primitives $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ denotes that $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

@@

Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types.

@@

Rule \textsc{TCheck} states that checked properties $\xcheckP{\PStatus}{e}$ require a boolean property $e$.

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee must be boolean expressions, and the guarantee can refer to the result as $x$.

@@

\begin{figure}[t]
  \begin{mathpar}
    \boxed{\bigstep{\Sigma}{e}{v}}
  \end{mathpar}
  \begin{mathpar}
    \ruleAx{\bigstep{\Sigma; \sigma}{x}{\sigma(x)}}{Var}
    \quad
    \ruleAx{\bigstep{\Sigma}{v}{v}}{Value}

@@

    \boxed{\bigstepalways{\Sigma}{e}}
  \end{mathpar}
  \begin{mathpar}
    \ruleAx{\bigsteps{\cdot}{e}{\cdot}}{$\mbox{Steps}_0$}

@@

At a high level, evaluation unfolds recursive streams to determine a value.
For example, to evaluate the earlier sum example with input $\textit{ints} = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumintsonetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream one step to get $(\xfby{0}{\sumrec}) + \textit{ints}$.
Evaluation of primitives is standard.
To evaluate variables, we look for the variable in the current (rightmost) heap:
$$
\ruleIN{
}{
  \bigstep{\sumintsonetwo}{\textit{ints}}{2}
}{Var}
$$

For delays, we discard the current heap and continue evaluation with the history prefix:
$$
\ruleIN{
  \bigstep{\sgl{\textit{ints} \mapsto 1}}{\sumrec}{1}
}{
  \bigstep{\sumintsonetwo}{\xfby{0}{\sumrec}}{1}
}{$\mbox{Fby}_S$}
$$

Returning to \autoref{f:core-bigstep},
rule {\sc Var} evalutes a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap.
Rules {\sc Value} and {\sc Let} are standard.
Rule {\sc Prim} evaluates a primitive $p$ applied to many arguments $e_1$ to $e_n$ by evaluating each argument separately; we then apply the primitive with prim-sem metafunction.

For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value.
When there is no previous value -- the streaming history only contains the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
Otherwise, rule $\textsc{Fby}_S$ applies; we evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.

>>>>

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
To verify programs, Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics.
To execute programs, Pipit can generate executable code, which is total and semantics-preserving.

@@

\begin{itemize}
  \item we motivate the need to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:core:causality}, \autoref{s:transition:proof});
  \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
  \item finally, we evaluate Pipit by implementing the high-level logic of a time-triggered Controller Area Network (CAN) bus driver and verifying an abstract model of a key operation (\autoref{s:evaluation}).
\end{itemize}

@@

To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered Controller Area Network (CAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.

@@

The schedule of our time-triggered driver is determined by a constant array of triggers, sorted by their associated time-mark.
The driver maintains an index that refers to the current trigger.
At each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
At each time step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.

@@

\subsection{The time-triggered system matrix}

The schedule of the time-triggered network is abstractly described by a \emph{system matrix}, consisting of rows of \emph{basic cycles}, columns of \emph{transmission columns}, and cells of optional messages.
Each basic cycle is identified by its cycle index and each transmission column has an associated time-mark.

\autoref{f:tt-systemmatrix-ok} (left) shows an example system matrix with cycles \tt{C0} and \tt{C1} and transmission columns at time-marks 0, 1 and 2.
For this example, we assume that one message can be sent per clock cycle. To execute this system matrix, we synchronise the local time to zero at the start of basic cycle \tt{C0}.
After a basic cycle completes, the nodes on the network synchronise before execution continues to the next basic cycle.

\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding configuration for the triggers array.
The enabled set denotes the basic cycles for which a trigger is active.

The system has strict timing requirements which restrict how triggers can be defined.
In this example, each trigger has a unique time; in general, trigger times can overlap, but they need to be enabled on distinct cycles.
Additionally, the schedule must allow sufficient time for the driver to skip over the disabled triggers.
Concretely, we could postpone trigger 1 to send message \tt{B} at time-mark 2, as triggers 1 and 2 have distinct cycles.
However, we could not bring forward trigger 2 to send message \tt{C} at time-mark 1: the driver can only process one trigger per tick, and it takes two steps to reach trigger 2 from the start of the array.

\begin{figure}
  \begin{minipage}{0.38\textwidth}
\begin{tabular}{r|ccc}
   & TM0 & TM1 & TM2 \\
  \hline
  C0 & MSG A & MSG B & - \\
  C1 & MSG A & - & MSG C
\end{tabular}
\end{minipage}
\begin{minipage}{0.6\textwidth}
\small
\begin{verbatim}
  0: { time_mark = 0; enabled = {C0,C1}; msg = A; }
  1: { time_mark = 1; enabled = {C0};    msg = B; }
  2: { time_mark = 2; enabled = {C1};    msg = C; }
\end{verbatim}
\end{minipage}

\caption{Left: system matrix; right: corresponding triggers array configuration}
\label{f:tt-systemmatrix-ok}
\end{figure}

We impose three restrictions on the triggers array: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.

@@

We can now implement the trigger-fetch logic, which keeps track of the current trigger.
The trigger-fetch logic uses the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.

\begin{tabbing}
  \tt{MM}\= \tt{MM} \= \tt{let index} \= \kill
  \tt{let} trigger_fetch ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
    \> $\tt{rec } \textit{index}.$ \\
    \> \> $\tt{let } \textit{inc} = \text{false} \tt{ fby } ((\text{time_mark}~\textit{index}) \le \textit{time} ~\vee~ \neg (\text{enabled}~\textit{index}~\textit{cycle})) \tt{ in}$\\
    \> \> $\tt{let } \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} \tt{ in}$ \\
    \> \> $\text{pose}_1~(\text{lemma_can_reach_next}~\textit{cycle})~\textit{index}$; \\
    \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
    \> \> $\textit{index}$
\end{tabbing}

@@

We use the $\textit{pose}_1$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it; the subscript 1 indicates that the lemma is being applied to one streaming argument (the index).
We then state an invariant as a deferred property.
Informally, the invariant states that, either the current active trigger is not late, or the next active trigger after the current index is in the future and we can reach it in time.

With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
The contract we state is that if the environment ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.

\begin{tabbing}
  \tt{MM}\= \tt{MM} \= \tt{MMMMMMMMMMMMM} \= \kill
  \tt{let} $\text{trigger_fetch}_{\PSValid}$ ($\textit{cycle}$: $\NN$): $\stream \NN \to \stream \NN$ = \\
  \> \tt{let} $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
  \> \> \tt{rely} = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} );$ \\
  \> \> \tt{guar} = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
  \> \> \> $\implies (\text{time_mark}~\textit{index}) \ge \textit{time});$ \\
  \> \> \tt{body} = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cycle}~\textit{time} );$ \\
  \> \} \tt{in} \\
  \> \tt{assert} $(\text{Contract.inductive_check}~\textit{contract})$ \tt{by} (pipit_simplify ()); \\
  \> $\text{Contract.stream_of_contract}_1~\textit{contract}$
\end{tabbing}

@@

When this function is used in other parts of the program, the caller must ensure that the environment satisfies the rely clause.
In the core language, this is tracked by another deferred property status attached to the contract; we will discuss this further in \autoref{s:core}.

\section{Core language}
\label{s:core}

\begin{figure}
  \includegraphics[width=\textwidth]{figures/core-structure.pdf}
\caption{Architecture of Pipit. The gray boxes and solid arrows are defined in this paper. The white boxes and dashed arrows are trusted components. The labels next to the arrows denote properties of the translation.
}
\label{f:core:structure}
\end{figure}

@@

\autoref{f:core:structure} shows the high-level architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.

@@

\begin{figure}
  \[
  \begin{array}{lrlr}
    e, e' & := & v ~|~ x ~|~ p(\ov{e}) & \mbox{(values, variables and operations)} \\
          & | & \xfby{v}{e} ~|~ \xrec{x}{e[x]} & \mbox{(delayed and recursive streams)} \\ & | & \xlet{x}{e}{e'[x]} & \mbox{(let-expressions)}\\
          & | & \xcheckP{\PStatus}{e_{\text{prop}}} & \mbox{(checked properties)} \\
          & | & \xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}{\eguar[x]}} & \mbox{(rely-guarantee contracts)} \\
\\
    v & := & n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ r \in \mathbb{R} ~|~ \hdots  & \mbox{(values)} \\
    p & := & (+) ~|~ (-) ~|~ (\times) ~|~ \tt{if-then-else} ~|~ \hdots & \mbox{(primitives)} \\
    \\
    \PStatus & := & \PSValid ~|~ \PSUnknown & \mbox{(property statuses: valid or unknown)}\\
    \\
    V & := & \cdot ~|~ V;v & \mbox{(streams of values)} \\
    \sigma & := & \sgl{\ov{x \mapsto v}} & \mbox{(heaps)} \\
    \Sigma & := & \cdot ~|~ \Sigma;\sigma & \mbox{(streaming history environments)} \\
\tau, \tau' & := & \mathbb{N} ~|~ \mathbb{B} ~|~ \tau \times \tau ~|~ \hdots & \mbox{(value types)} \\
    \Gamma & := & \cdot ~|~ x : \tau, \Gamma & \mbox{(type environments)}  \\
    \end{array}
  \]
  \caption{Pipit core language grammar, which contains expressions $e$, values $v$, primitive operations $p$, and property statuses $\PStatus$.}
  \label{f:core-grammar}
\end{figure} 
The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

@@

Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ involve two verification conditions.
Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
Conceptually, then, a contract could have two property statuses: one for the definition and one for the instantiation.
However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.

For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.

Streams $V$ are represented as a sequence of values; streaming history environments $\Sigma$ are streams of heaps.
Types $\tau$ and type environments $\Gamma$ are standard.

@@

  \caption{Typing rules for Pipit; the judgment $\typing{\Gamma}{e}{\tau}$ denotes that expression $e$ describes a \emph{stream} of values of type $\tau$. Two auxiliary functions are used for values and primitive operations; their definitions are standard and are omitted.}\label{f:core-typing}
\end{figure} 
We define the typing judgments for Pipit in \autoref{f:core-typing}.
Most of the typing rules are standard for an unclocked Lustre.
The typing judgment $\typing{\Gamma}{e}{\tau}$ denotes that, in an environment of streams $\Gamma$, expression $e$ denotes a stream of type $\tau$.
This core typing judgment differs from the surface syntax used in \autoref{s:motivation}, which used an explicit stream type; for the core language, we instead assume that everything is a stream.

For values, we use an auxiliary definition $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
Likewise, for primitives we use $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

@@

Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types; the result is the same type.

@@

Rule \textsc{TCheck} states that statically checking a property $\xcheckP{\PStatus}{e}$ requires a boolean property $e$ and returns unit.

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of some type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee clauses must be boolean expressions.
Additionally, the guarantee clause can refer to the result value as $x$.

@@

\begin{figure}
  \begin{mathpar}
    \boxed{\bigstep{\Sigma}{e}{v}}
  \end{mathpar}

  \begin{mathpar}
    \ruleAx{\bigstep{\Sigma}{v}{v}}{Value}
    \quad
    \ruleAx{\bigstep{\Sigma; \sigma}{x}{\sigma(x)}}{Var}

@@

    \boxed{\bigstepalways{\Sigma}{e}}
  \end{mathpar}

  \begin{mathpar}
    \ruleAx{\bigsteps{\cdot}{e}{\cdot}}{$\mbox{Steps}_0$}

@@

Rule {\sc Value} states that evaluating a value results in the value itself.

Rule {\sc Var} states that to evalute a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap, we look up the variable in $\sigma$.

Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.

Rule $\textsc{Fby}_1$ evaluates a followed-by expression when the streaming history contains only a single element.
Here, $\xfby{v}{e}$ evaluates to $v$, as there is no previous value of $e$ to use.

Rule $\textsc{Fby}_S$ evaluates a followed-by expression when the streaming history contains multiple entries.
In this case, $\xfby{v}{e}$ evaluates the previous value of $e$ by discarding the most recent entry from the streaming history.
